Analyzing DECREASE trials to estimate evidence of data manipulation

### 2017-07-27 submission/manuscript.Rmd
Revision round gazilion

---
title: "Analyzing DECREASE trials to estimate evidence of data manipulation"
author: "Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper, Markus Hollman"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document: default
csl: ../bibliography/ama.csl
bibliography: ../bibliography/library.bib
---

@@

@@

The trials subject to concerns regarding scientific integrity were the Dutch DECREASE-I and DECREASE-IV trials [@commissie2011;@commissie2012;@commissie2013]. The committees that investigated the integrity of the DECREASE trials reported that data manipulation was likely but that the extent of the data manipulation remained unclear [@commissie2011;@commissie2012;@commissie2013]. Moreover, the latest guidelines still recommend the usage of beta-blockers in the perioperative period in certain cases [@esc2014;@Fleisher_2014], where some of these guidelines are based on other work by the PI of the DECREASE trials. Considering the potential harmful consequences of guidelines based on work by someone who has been repeatedly investigated for breaching scientific integrity, we aim to estimate the extent of data manipulation in the DECREASE studies [@poldermans1999;@dunkelgrun2009;@commissie2011;@commissie2012;@commissie2013] to further stimulate the debate on using beta-blockers in the perioperative period patients undergoing non-cardiac surgery.

The reports on the integrity of the DECREASE trials primarily focused on the provenance of the raw data but did not investigate the extent to which the DECREASE trials deviated from comparable trials. Provenance is primarily concerned with the origins of the data, verifying things such as (but not limited to) the informed consent and whether data corresponded to patient files. The committee reports did not neglect statistical evaluation however: according to the report a statistical expert evaluated the applicability of forensic statistical methods [@commissie2013] to evaluate results of trials separately (i.e., DECREASE-I, DECREASE-IV) although it lacks details as to how this evaluation took place. The expert concluded that the previously applied methods by him were not applicable for use in this case. Nonetheless, in this report we compare across trials, which is a method that has previously been used to monitor trial data quality or to test for potential data anomalies [@Buyse1999-jq;@10.1111/anae.13938]. Moreover, comparing across trials instead of evaluating them separately  has previously proven to be effective in detecting data manipulation [@Knepper2016-la]. Comparing the DECREASE trials to other published trials studying the effectiveness of beta-blockers with respect to perioperative mortality could prove informative of the potential extent of the manipulation in the DECREASE trials.

@@

To statistically investigate the evidence for data manipulation in the DECREASE studies [@dunkelgrun2009;@poldermans1999], we took three steps. First, we reproduced the findings from the 2014 meta-analysis by Bouri et al. [@bouri2014], which contained sufficient information to estimate the deviation of the DECREASE trials from other published trials on beta-blockers. We also included type of beta-blocker to inspect whether this is predictive of the effect of beta-blockers on perioperative mortality. Second, we evaluated the probability that the DECREASE trials occurred assuming no data manipulation. Third, we reversed this assumption and assumed that data manipulation did occur and estimated how many data points would have to be manipulated in order to reproduce the results of the DECREASE trials, if the other published trials are regarded as estimating the true effect of beta-blockers on perioperative mortality in patients undergoing non-cardiac surgery. Considering the committees investigating the scientific integrity of the DECREASE trials was unable to assess this, we consider it worthwhile to investigate this further.

@@

We computed the log risk ratio (i.e., log RR) for each study and pooled these using the `R` package `metafor` [@viechtbauer2010]. We estimated a weighted random-effects model using the restricted maximum-likelihood estimator (i.e., `REML`) [@viechtbauer2005] to estimate the variance of effects. We used the default weighting procedure in the `metafor` package. We added 0.5 to each cell count, as is common in meta-analyses on risk- and odds ratios in order to prevent computational artefacts [@agresti2002; p397]. The 2014 meta-analysis [@bouri2014] did not specify the variance estimate used; hence, minor discrepancies between our estimates and the original estimates could be due to differences in the estimation procedure.

@@

We were able to closely reproduce the estimates for the different sets of studies (Figure 2 of the 2014 meta-analysis [@bouri2014]). Bouri et al. differentiated between the estimates from the non-DECREASE trials ($k=`r sum(bouri_data$decrease == 0)`$) and the DECREASE trials ($k=`r sum(bouri_data$decrease == 1)`$). We confirmed the effect size estimates and the variance estimates for both the non-DECREASE- and the DECREASE trials, except for some minor discrepancies at the second decimal level. Table `r tabnr` depicts the original and reproduced values for both sets of studies.

```{r echo=FALSE,results='asis'}
table1 <- matrix('', ncol = 5, nrow = 5)
table1[1,3] <- 'Risk ratio'
table1[1,4] <- '$\\tau^2$'
table1[1,5] <- 'Confidence interval'
table1[2,1] <- 'Non-DECREASE (k=9)'
table1[4,1] <- 'DECREASE (k=2)'
table1[c(2,4),2] <- 'Original'
table1[c(3,5),2] <- 'Reproduced'
table1[2,3] <- 1.27
table1[2,4] <- 0.00
table1[2,5] <- '[1.01; 1.60]'

table1[3,3] <- round(exp(mod_clean$b), 2)
table1[3,4] <- round(mod_clean$tau2, 2)
table1[3,5] <- sprintf('[%s; %s]', round(exp(mod_clean$ci.lb), 2), round(exp(mod_clean$ci.ub), 2))

table1[4,3] <- 0.42
table1[4,4] <- 0.29
table1[4,5] <- '[0.15; 1.23]'

table1[5,3] <- round(exp(mod_dirty$b), 2)
table1[5,4] <- round(mod_dirty$tau2, 2)
table1[5,5] <- sprintf('[%s; %s]', round(exp(mod_dirty$ci.lb), 2), round(exp(mod_dirty$ci.ub), 2))

@@

Second, we meta-analyzed all studies combined, including a dummy predictor for the DECREASE and non-DECREASE studies to reproduce results presented in Figure 4 of the 2014 meta-analysis [@bouri2014]. Surprisingly, our results showed a bit more evidence against equal subgroups than the original meta-analysis [@bouri2014] (original: $\chi^2(1)=3.91,p=.05$; reproduced: $\chi^2(1)=`r round(mod_mods$QM, 2)`,p=`r round(mod_mods$QMp, 3)`$). Additionally, the original analyses showed substantial residual heterogeneity ($I^2=74.4$\%) whereas we found no residual heterogeneity ($I^2=`r round(i2_mods, 3)`$\%). Different variance estimates (e.g., `DerSimonian-Laird` instead of `REML`) did not resolve this difference. We tried to clarify these discrepancies by e-mailing the original authors (including a reminder after several weeks) but did not receive a response. Nonetheless, the broad strokes of the meta-regression confirmed that the DECREASE trials were the determining predictor for the effectiveness of beta-blockers (including DECREASE: $RR=`r round(exp(mod_mods$b[1] + mod_mods$b[2]), 3)`$; excluding DECREASE: $RR=`r round(exp(mod_mods$b[1]), 3)`$).
<!--- Emailed bouri about this on 2016 03 16 -->

@@

Additionally, and exploratively, we evaluated the predictive effect of the type of beta-blocker used in the trials. The DECREASE trials remained predictive of decreased mortality ($RR=`r round(exp(mod_mods_type$b[1] + mod_mods_type$b[2] + mod_mods_type$b[3]), 3)`$), <!-- DECREASE only uses bisoprolol! -->
whereas the non-DECREASE trials provide tentative, but uncertain, evidence that atenolol results in lower mortality ($RR=`r round(exp(mod_mods_type$b[1]), 3)`$). Nonetheless, for other beta-blockers in the non-DECREASE trials, there is still tentative and uncertain evidence that beta-blockers increase mortality (bisoprolol: $RR=`r round(exp(mod_mods_type$b[1] + mod_mods_type$b[3]), 3)`$; metoprolol: $RR=`r round(exp(mod_mods_type$b[1] + mod_mods_type$b[4]), 3)`$; propranolol: $RR=`r round(exp(mod_mods_type$b[1] + mod_mods_type$b[5]), 3)`$). Table `r tabnr` shows the meta-regression results in full.

@@

Based on the non-DECREASE estimates from Step 1, we estimated the probability of obtaining the results in the DECREASE trials. To this end, we assumed that the non-DECREASE trials provide a valid representation of the true effect of beta-blockers on perioperative mortality (similar to Bouri et al. [@bouri2014]). The estimated probability of the observed data under a given true effect is also known as the veracity of the data [@peters2015]. We assumed that the non-DECREASE studies estimated the true effect distribution of perioperative beta-blockade on mortality, not perturbed by publication bias due to statistical (non)significance. Publication bias was assumed to not be a problem because a substantial number of nonsignificant effects are included in the dataset (9 of 11 results are nonsignificant).

@@

```{r, echo = FALSE}
mod_clean_pred <- predict(mod_clean)
mod_dirty_pred <- predict(mod_dirty)
```

Based on the estimated mean log RR and its credible interval in the non-DECREASE studies, we computed the probability of the observed log RR in the DECREASE trials. The estimates of the non-DECREASE studies were obtained from Step 1, which include the estimated log RR (i.e., `r round(mod_clean$b, 2)`), and its 95\% credibility interval as provided by the package `metafor` (i.e., `r sprintf('[%s; %s]', round(mod_clean_pred$cr.lb, 3), round(mod_clean_pred$cr.ub, 3))`). The meta-analysis model assumes a normal distribution of population effects with the estimated effect as the mean of the distribution. The 95\% credibility interval denotes the bounds of the normal distribution that covers 95\% of the density, where the standard deviation is calculated as the distance from the mean to either bound, divided by 1.96. This allows for an approximation of the population effect distribution, as depicted in Figure `r fignr`.

```{r, echo = FALSE}
temp <- seq(-3, 3, .001)

pdf(sprintf('%sfigures/fig1.pdf', x12), width = 8, height = 6)
par(mar = c(5,0,0,0))
plot(x = temp,
     dnorm(x = temp,
           mod_clean$b,
           (mod_clean_pred$pred - mod_clean_pred$cr.lb)/1.96),
     type = 'l',
     bty = 'n',
     # xaxt = 'n',
     yaxt = 'n',
     xlab = 'log(RR)',
     ylab = '')

points(x = bouri_data$ln_rr[bouri_data$decrease == 1][1], y = 0)
text(x = bouri_data$ln_rr[bouri_data$decrease == 1][1],
     y = .15,
     labels = 'DECREASE-I',
     cex = .8)

points(x = bouri_data$ln_rr[bouri_data$decrease == 1][2], y = 0)
text(x = bouri_data$ln_rr[bouri_data$decrease == 1][2] - .05,
     y = .15,
     labels = 'DECREASE-IV',
     cex = .8)

invisible(dev.off())
```

```{r figure 1, fig.cap = 'Density plot of the estimated true effect distribution based on the non-DECREASE studies only, with the position of the DECREASE studies highlighted.', out.width = '80%', fig.asp = .5, fig.align = 'center', echo = FALSE}
knitr::include_graphics(sprintf("%sfigures/fig1.pdf", x12))
fignr <- fignr + 1
```

Based on the estimated effect distribution from the non-DECREASE trials, we calculated the probability of each DECREASE trial result, or a more extreme result. In other words, we computed the $p$-value for the null hypothesis that the DECREASE trials arise from the same effect distribution as the non-DECREASE trials. This assumes that the information available from the other trials is informative of the true population effect.

@@

Figure `r fignr` indicates that the DECREASE trials are highly unlikely under the estimated effect distribution based on the non-DECREASE trials. More specifically, the results from DECREASE-I (or more extreme) have a probability of $`r prob[1]`$ (approximately 7 in a [quattuordecillion](https://en.wikipedia.org/wiki/Names_of_large_numbers); $z=`r round(zval[1], 3)`$) and the results from DECREASE-IV have a probability of $`r prob[2]`$ (approximately 7 in a billion; $z=`r round(zval[2], 3)`$). This indicates that the DECREASE trial results are unlikely to have come from the same population effect distribution as the non-DECREASE trials. Moreover, observing two of such extremely unlikely results jointly, as in the DECREASE trials, is nearly impossible, $`r prob[1]*prob[2]`$. Hence, this result indicates that the DECREASE trials are severely different from the non-DECREASE trials.

Results from Step 1 indicated that no between-trial variance (i.e., homogeneity; $\tau^2=0$) of the effects was observed; given the small number of trials included (i.e., `r mod_clean$k`) this estimate is highly uncertain, however. The total $N$ across the non-DECREASE trials was `r sum(bouri_data$control_n[bouri_data$decrease == 0] + bouri_data$beta_n[bouri_data$decrease == 0])`.We conducted sensitivity analyses to see how dependent results are on the heterogeneity estimate (not preregistered; [osf.io/vnmzc](https://osf.io/vnmzc)). The probability of observing the DECREASE trials stays approximately below 1 out of 1000 until the variance estimate is 0.25; the probability stays approximately below 1 out of 100 until the variance estimate is 0.43 (see Figure `r fignr`). To put these numbers into context, a variance of 0.25 would suggest that results of perioperative beta-blockade vary substantially due to contextual circumstances of the study, even if perioperative beta-blockade has no effect whatsoever (RRs between `r round(exp(0 - .25), 3)` and `r round(exp(0 + .25), 3)` in ~64\% of the cases).

@@

for (t2 in tau2){
  mod_clean_sens <- metafor::rma(yi = ln_rr,
                                 sei = se_ln_rr, tau2 = t2,
                                 method = "REML",
                                 data = bouri_data[bouri_data$decrease == 0, ],
                                 weighted = TRUE)
  mod_clean_pred_sens <- predict(mod_clean_sens)
  prob_sens <- pnorm(q = bouri_data$ln_rr[bouri_data$decrease == 1],
                     mod_clean_pred_sens$pred,
                     (mod_clean_pred_sens$pred - mod_clean_pred_sens$cr.lb)/1.96) * 2
  prob_sens_t[i] <- prob_sens[1] * prob_sens[2]

@@

par(mar = c(4, 4, 0, 0))
plot(tau2, prob_sens_t, type = 'l', xlab = TeX('$\\tau^2$'),
     ylab = 'p-value', bty = 'n')
points(x = tau2[c(26, 44)], y = prob_sens_t[c(26, 44)])
text(x = tau2[c(26, 44)],
     y = prob_sens_t[c(26, 44)] + .003,
     labels = c(TeX('$\\tau^2=.25$'), TeX('$\\tau^2=.43$')),
     cex = .8)
invisible(dev.off())
```
```{r figure 2, fig.cap = 'Sensitivity analyses for the p-value that indicates the probability of observing the results from the DECREASE studies, or more extreme results, based on the estimated true effect (non-DECREASE trials) and the accompanying variance estimate.', out.width = '80%', fig.asp = .5, fig.align = 'center', echo = FALSE}
knitr::include_graphics(sprintf("%sfigures/fig2.pdf", x12))
fignr <- fignr + 1
```

@@

```{r echo = FALSE}
# This is for easy reference
# And to easily include in text
iterations <- 10000
```
We estimated the number of data points that would need to be manipulated to arrive at the estimates from the DECREASE trials, given that the non-DECREASE trials represent the true effect of perioperative beta-blockade. In contrast to Step 2, which assumes no data manipulation occurred, Step 3 assumes that the DECREASE trials might in fact contain manipulated data. The estimates from Step 3 provide an indication of the extent of potential data manipulation in the DECREASE studies [@commissie2011;@commissie2012;@commissie2013;@bouri2014].

@@

In order to estimate the number of manipulated data points, we first estimated the probability of perioperative mortality (in log odds) in each trial arm for each trail stratum. As such, we estimate mortality odds four times: once per condition (beta-blocker or control) per trial type (DECREASE- and non-DECREASE trials). For all combinations of condition and trial type, we ran a meta-analysis applying similar methods used in Step 1, resulting in four meta-analytic absolute mortality estimates with corresponding effect variances (see Table 3; one estimate per cell). Throughout the simulations, we used the point estimates (i.e., fixed effect) to simulate genuine- and manipulated data, but supplemented this by using the more uncertain distribution estimates (i.e., random effects) as sensitivity analyses.

```{r echo=FALSE, results='asis'}
table2 <- matrix('', ncol = 3, nrow = 3)
table2[1,1] <- ''
table2[1,2] <- 'Dead'
table2[1,3] <- 'Alive'

>>>>

---
title: "Analyzing DECREASE trials to estimate evidence of data manipulation"
author: "[Chris H. J. Hartgerink](https://orcid.org/0000-0003-1050-6809), E. M. Kemper, [Markus W. Hollman](https://orcid.org/0000-0001-8248-0244), [Gerben ter Riet](https://orcid.org/0000-0002-2231-7637)"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document: default
csl: ../bibliography/ama.csl
bibliography: ../bibliography/library.bib
---

# Abstract

The effect of beta-blockers on perioperative mortality in non-cardiac surgery has been controversial due to concerns regarding the scientific integrity of the DECREASE-I and DECREASE-IV trials. In this report, we investigate these DECREASE trials and model their veracity (i.e., the probability that these effects or more extreme occurred naturally) and estimate how many data points might have been manipulated in the DECREASE trials. Our research indicates that the DECREASE trials are nearly impossible if we assume they investigate the same effect as the non-DECREASE trials and under that assumption, our results provide evidence that at least some data points were manipulated. However, these differences might also be due to different conceptual approaches as to how beta-blockade might prevent mortality in non-cardiac surgery. Considering this, we recommend new and more extensively controlled, confirmatory trials to determine whether there is any use in administering beta-blockers in order to decrease perioperative mortality.

# Author note

All manuscript materials are available at https://github.com/chartgerink/2015poldermans and are preserved at Zenodo [@].

@@

# Introduction

@@

The trials subject to concerns regarding scientific integrity were the Dutch DECREASE-I and DECREASE-IV trials [@commissie2011;@commissie2012;@commissie2013]. The committees that investigated the integrity of the DECREASE trials reported that data manipulation was likely but that the extent of the data manipulation remained unclear [@commissie2011;@commissie2012;@commissie2013]. Moreover, the latest guidelines still recommend the usage of beta-blockers in the perioperative period in certain cases [@esc2014;@Fleisher_2014], where some of these guidelines are based on other work by the PI of the DECREASE trials. Considering the potential harmful consequences of guidelines based on work by someone who has been repeatedly investigated for breaching scientific integrity, we aim to estimate the extent of data manipulation in the DECREASE studies [@poldermans1999;@dunkelgrun2009;@commissie2011;@commissie2012;@commissie2013] to further stimulate the debate on using beta-blockers in the perioperative period for patients undergoing non-cardiac surgery.

The reports on the integrity of the DECREASE trials primarily focused on the provenance of the raw data but did not investigate the extent to which the DECREASE trials deviated from comparable trials. Provenance is primarily concerned with the origins of the data, verifying things such as (but not limited to) the informed consent and whether data corresponded to patient files. The committee reports did not neglect statistical evaluation however: according to the report a statistical expert evaluated the applicability of forensic statistical methods [@commissie2013] to evaluate results of trials separately (i.e., DECREASE-I, DECREASE-IV) although the report lacks details as to how this evaluation took place. The expert concluded that the previously applied methods by him were not applicable for use in this case. Nonetheless, in this report we compare across trials, which is a method that has previously been used to monitor trial data quality or to test for potential data anomalies [@Buyse1999-jq;@10.1111/anae.13938]. Moreover, comparing across trials instead of evaluating them separately has previously proven to be effective in detecting data manipulation [@Knepper2016-la]. Comparing the DECREASE trials to other published trials studying the effectiveness of beta-blockers with respect to perioperative mortality could prove informative of the potential extent of the manipulation in the DECREASE trials.

@@

To statistically investigate the evidence for data manipulation in the DECREASE studies [@dunkelgrun2009;@poldermans1999], we took three steps. 
First, we reproduced the findings from the 2014 meta-analysis by Bouri et al. [@bouri2014], which contained sufficient information to estimate the deviation of the DECREASE trials from other published trials on beta-blockers. We also included type of beta-blocker to inspect whether this is predictive of the effect of beta-blockers on perioperative mortality. 
Second, we evaluated the probability that the DECREASE trials (or more extreme effects) arose from the same effect distribution as the non-DECREASE trials, which are assumed to be the true effect of beta-blockers on perioperative mortality in patients undergoing non-cardiac surgery. 
Third, we estimated how many data points would have to be manipulated in order to reproduce the results of the DECREASE trials if the initial, non-manipulated results arose from the effect estimates as obtained from the non-DECREASE trials. Considering the committees investigating the scientific integrity of the DECREASE trials was unable to assess this, we consider it worthwhile to investigate this further.

@@

We computed the log risk ratio (i.e., log RR) for each study and pooled these using `v2.0.0` of the `R` package `metafor` [@viechtbauer2010]. We estimated a weighted random-effects model using the restricted maximum-likelihood estimator (i.e., `REML`) [@viechtbauer2005] to estimate the variance of effects. We used the default weighting procedure in the `metafor` package. We added 0.5 to each cell count, as is common in meta-analyses on risk- and odds ratios in order to prevent computational artefacts [@agresti2002]. <!--p397-->
The 2014 meta-analysis [@bouri2014] did not specify the variance estimate used, hence, (minor) discrepancies between our estimates and the original estimates could be due to differences in the estimation procedure.

@@

We were able to closely reproduce the estimates for the different sets of studies (Figure 2 of the 2014 meta-analysis [@bouri2014]). Bouri et al. differentiated between the estimates from the non-DECREASE trials ($k=`r sum(bouri_data$decrease == 0)`$) and the DECREASE trials ($k=`r sum(bouri_data$decrease == 1)`$). We confirmed the effect size estimates and the variance estimates for both the non-DECREASE- and the DECREASE trials, except for some discrepancies at the second decimal level for the estimated effect sizes and a somewhat larger difference between the variance estimates of the DECREASE studies. Table `r tabnr` depicts the original and reproduced values for both sets of studies.

```{r echo=FALSE,results='asis'}
table1 <- matrix('', ncol = 5, nrow = 5)
table1[1,3] <- 'Risk ratio'
table1[1,5] <- '$\\tau^2$'
table1[1,4] <- 'RR Confidence interval'
table1[2,1] <- 'Non-DECREASE (k=9)'
table1[4,1] <- 'DECREASE (k=2)'
table1[c(2,4),2] <- 'Original'
table1[c(3,5),2] <- 'Reproduced'
table1[2,3] <- 1.27
table1[2,5] <- 0.00
table1[2,4] <- '[1.01; 1.60]'

table1[3,3] <- round(exp(mod_clean$b), 2)
table1[3,5] <- round(mod_clean$tau2, 2)
table1[3,4] <- sprintf('[%s; %s]', round(exp(mod_clean$ci.lb), 2), round(exp(mod_clean$ci.ub), 2))

table1[4,3] <- 0.42
table1[4,5] <- 0.29
table1[4,4] <- '[0.15; 1.23]'

table1[5,3] <- round(exp(mod_dirty$b), 2)
table1[5,5] <- round(mod_dirty$tau2, 2)
table1[5,4] <- sprintf('[%s; %s]', round(exp(mod_dirty$ci.lb), 2), round(exp(mod_dirty$ci.ub), 2))

@@

Second, we meta-analyzed all studies combined, including a dummy predictor for the DECREASE and non-DECREASE studies to reproduce results presented in Figure 4 of the 2014 meta-analysis [@bouri2014]. Our results showed a bit more evidence against equal subgroups than the original meta-analysis [@bouri2014] (original: $\chi^2(1)=3.91,p=.05$; reproduced: $\chi^2(1)=`r round(mod_mods$QM, 2)`,p=`r round(mod_mods$QMp, 3)`$). Additionally, the original analyses showed substantial residual heterogeneity ($I^2=74.4$\%) whereas we found no residual heterogeneity ($I^2=`r round(i2_mods, 3)`$\%). Different variance estimates (e.g., `DerSimonian-Laird` instead of `REML`) did not resolve this difference. We tried to clarify these discrepancies by e-mailing the original authors (including a reminder after several weeks) but did not receive a response. Nonetheless, the broad strokes of the meta-regression confirmed that the DECREASE trials were the determining predictor for the effectiveness of beta-blockers (including DECREASE: $RR=`r round(exp(mod_mods$b[1] + mod_mods$b[2]), 3)`$; excluding DECREASE: $RR=`r round(exp(mod_mods$b[1]), 3)`$).
<!--- Emailed bouri about this on 2016 03 16 -->

@@

Additionally, and exploratively, we evaluated the predictive effect of the type of beta-blocker used in the trials. Descriptively, the DECREASE trials remained predictive of decreased mortality ($RR=`r round(exp(mod_mods_type$b[1] + mod_mods_type$b[2] + mod_mods_type$b[3]), 3)`$), <!-- DECREASE only uses bisoprolol! -->
whereas the non-DECREASE trials provide tentative evidence that atenolol results in lower mortality ($RR=`r round(exp(mod_mods_type$b[1]), 3)`$). Nonetheless, for other beta-blockers in the non-DECREASE trials, there is descriptive evidence that beta-blockers could increase mortality (bisoprolol: $RR=`r round(exp(mod_mods_type$b[1] + mod_mods_type$b[3]), 3)`$; metoprolol: $RR=`r round(exp(mod_mods_type$b[1] + mod_mods_type$b[4]), 3)`$; propranolol: $RR=`r round(exp(mod_mods_type$b[1] + mod_mods_type$b[5]), 3)`$). Table `r tabnr` shows the meta-regression results in full. We do note that the DECREASE studies only use bisoprolol and any estimates for other beta-blockers are extrapolations.

@@

Based on the effect estimates for the non-DECREASE trials from Step 1, we estimated the probability that the observed effects from the DECREASE studies (or more extreme) occurred naturally. We assumed that the non-DECREASE studies estimated the true effect distribution of perioperative beta-blockade on mortality, not perturbed by publication bias due to statistical (non)significance. Publication bias was assumed to not be a problem because a substantial number of nonsignificant effects are included in the dataset (9 of 11 results are nonsignificant). Based on this effect distribution, we estimated the veracity of the DECREASE trials separately, which is the estimated probability of the observed data under a given true effect [@peters2015].

@@

```{r, echo = FALSE}
mod_clean_pred <- predict(mod_clean)
mod_dirty_pred <- predict(mod_dirty)

var1 <- mod_clean_pred$se^2 / sum(bouri_data$decrease == 0)
var_decrI <- (bouri_data$se_ln_rr[bouri_data$decrease == 1][1]^2 / sum(bouri_data$decrease == 0))
var_decrIV <- (bouri_data$se_ln_rr[bouri_data$decrease == 1][2]^2 / sum(bouri_data$decrease == 0))
decrI <- bouri_data$ln_rr[bouri_data$decrease == 1][1]
decrIV <- bouri_data$ln_rr[bouri_data$decrease == 1][2]
mu_ma <- mod_clean_pred$pred

df_I <- 1 + 9 - 2
t_I <- (decrI - mu_ma) / sqrt(var1 + var_decrI)
p_I <- pt(abs(t_I), df = df_I, lower.tail = FALSE) * 2

df_IV <- 1 + 9 - 2
t_IV <- (decrIV - mu_ma) / sqrt(var1 + var_decrIV)
p_IV <- pt(abs(t_IV), df = df_IV, lower.tail = FALSE) * 2
```

Based on the estimated effect distribution from the non-DECREASE trials, we calculated the probability of each DECREASE trial result, or a more extreme result. In other words, we computed the two-tailed $p$-value for the null hypothesis that the DECREASE trials arose from the same effect distribution as the non-DECREASE trials ($H_0:\mu_{\bar{x}_1-\bar{x}_2}=0$). To this end, we applied a Welch t-test [@doi:10.1093/biomet/34.1-2.28]. 
As means we used the observed log RR for the DECREASE trials (i.e., DECREASE-I: `r round(decrI, 3)`; DECREASE-IV: `r round(decrIV, 3)`) and the meta-analyzed log RR for the non-DECREASE trials (i.e., `r round(mu_ma, 3)`). 
As standard deviations we used the standard error for the DECREASE trials (i.e., DECREASE-I: `r round(var_decrI, 3)`; DECREASE-IV: `r round(var_decrIV, 3)`) and the standard error of the estimated log RR for the non-DECREASE trials (i.e., `r round(var1, 3)`). We initially preregistered that the DECREASE trials would be regarded as fixed in the computation of the veracity, which was erroneous because these also have their own standard error; hence, we applied the Welch test to take into account the uncertainty in the estimates of both the DECREASE and non-DECREASE trials.

@@

Figure `r fignr` indicates that the DECREASE trials are highly unlikely under the estimated effect distribution based on the non-DECREASE trials. More specifically, the results from DECREASE-I (or more extreme) have a probability of 1 in 10 000 ($t(`r df_I`)=`r round(t_I, 3)`,p=`r p_I`$) and the results from DECREASE-IV (or more extreme) have a probability of 1 in 1000 ($t(`r df_IV`)=`r round(t_IV, 3)`,p=`r p_IV`$). This indicates that the DECREASE trial results are unlikely to have come from the same population effect distribution as the non-DECREASE trials. Moreover, observing two of such extremely unlikely results jointly, as in the DECREASE trials, would occur in only 2 out of 10 million sets of two trials (i.e., $`r p_I*p_IV`$) according to this model. Hence, this result indicates that the DECREASE trials are severely different from the non-DECREASE trials.

Results from Step 1 indicated that no between-trial variance (i.e., homogeneity; $\tau^2=0$) of the effects was observed; given the small number of trials included (i.e., `r mod_clean$k`) this estimate is highly uncertain, however. The total $N$ across the non-DECREASE trials was `r sum(bouri_data$control_n[bouri_data$decrease == 0] + bouri_data$beta_n[bouri_data$decrease == 0])`. We conducted sensitivity analyses to see how dependent results are on the heterogeneity estimate (not preregistered; [osf.io/vnmzc](https://osf.io/vnmzc)). Fixing the variance estimate $\tau^2$ to .5, indicates that the probability of observing the DECREASE trials jointly is approximately 1.2 out of 100 000 (see Figure `r fignr`). To put these numbers into context, a variance of 0.25 would suggest that results of perioperative beta-blockade vary substantially due to contextual circumstances of the study, even if perioperative beta-blockade has no effect whatsoever (RRs between `r round(exp(0 - .25), 3)` and `r round(exp(0 + .25), 3)` in ~64\% of the cases).

@@

for (t2 in tau2){
  mod_clean_sens <- metafor::rma(yi = ln_rr,
                                 sei = se_ln_rr, tau2 = t2,
                                 method = "REML",
                                 data = bouri_data[bouri_data$decrease == 0, ],
                                 weighted = TRUE)
  mod_clean_pred_sens <- predict(mod_clean_sens)

  var1 <- mod_clean_pred_sens$se^2 / sum(bouri_data$decrease == 0)
  mu_ma <- mod_clean_pred_sens$pred

  df_I <- 1 + 9 - 2
  t_I <- (decrI - mu_ma) / sqrt(var1 + var_decrI)
  p_I <- pt(abs(t_I), df = df_I, lower.tail = FALSE) * 2

  df_IV <- 1 + 9 - 2
  t_IV <- (decrIV - mu_ma) / sqrt(var1 + var_decrIV)
  p_IV <- pt(abs(t_IV), df = df_IV, lower.tail = FALSE) * 2
  prob_sens_t[i] <- p_I* p_IV

@@

par(mar = c(4, 4, 1, 0))
plot(tau2, prob_sens_t, type = 'l', xlab = TeX('$\\tau^2$'),
     ylab = 'p-value', bty = 'n')
# points(x = tau2[c(26, 44)], y = prob_sens_t[c(26, 44)])
# text(x = tau2[c(26, 44)],
#      y = prob_sens_t[c(26, 44)] + .003,
#      labels = c(TeX('$\\tau^2=.25$'), TeX('$\\tau^2=.43$')),
#      cex = .8)
invisible(dev.off())
```
```{r figure 2, fig.cap = 'Sensitivity analyses for the p-value that indicates the probability of observing the results from the DECREASE studies, or more extreme results, based on the estimated true effect (non-DECREASE trials) and the accompanying variance estimate.', out.width = '80%', fig.asp = .5, fig.align = 'center', echo = FALSE}
knitr::include_graphics(sprintf("%sfigures/fig2.pdf", x12))
fignr <- fignr + 1
```

@@

```{r echo = FALSE}
# This is for easy reference
# And to easily include in text
iterations <- 10000
```
We estimated the number of data points that would need to be manipulated to arrive at the estimates from the DECREASE trials, given that the non-DECREASE trials represent the true effect of perioperative beta-blockade. In contrast to Step 2, which assumes no data manipulation occurred and that the DECREASE trials occurred naturally from the same effect distribution as the non-DECREASE trials, Step 3 assumes that the DECREASE trials might in fact contain manipulated data. The estimates from Step 3 provide an indication of the extent of potential data manipulation in the DECREASE studies [@commissie2011;@commissie2012;@commissie2013;@bouri2014].

@@

In order to estimate the number of manipulated data points, we first estimated the probability of perioperative mortality (in log odds) in each trial arm for each trail stratum. As such, we estimate mortality odds four times: once per condition (beta-blocker or control) per trial stratum (DECREASE- and non-DECREASE trials). For all four combinations of condition and trial type, we ran a meta-analysis applying similar methods used in Step 1, resulting in four meta-analytic absolute mortality estimates with corresponding effect variances. Throughout the simulations, we used the point estimates (i.e., fixed effect) to simulate genuine- and manipulated data, but supplemented this by using distribution estimates (i.e., random effects) as sensitivity analyses.

We applied the inversion method to estimate the number of manipulated data points in the DECREASE trials [@casella2002]. We assumed that if data are manipulated, each data point is manipulated in the same way and to the same extent. The inversion method iteratively hypothesizes that *X* out of *N* data points were manipulated (i.e., $X={0, 1, ..., N}$), assuming they were manipulated in the same way. For each combination of *X* and trial, we simulated `r iterations` datasets. Each simulated dataset contained *X* manipulated data points and *N-X* genuine data points. For each simulated dataset (exact simulation procedure in the next paragraph), we determined the likelihood of the results with 
\begin{equation}
\label{eq1}
L(\theta|\pi_{E},\pi_{C})=\pi_{E}^{n_{11}}(1-\pi_{E})^{n_{12}} \times \pi_{C}^{n_{21}}(1 - \pi_{C})^{n_{22}}
\end{equation}
where $\pi_{E}$ indicates the mortality rate in the beta-blocker condition as drawn from the meta-analytic effect distribution ($\pi_{C}$ indicates the mortality rate in the control condition). We estimated those parameters using the meta-analytic procedure described in the previous paragraph, resulting in the estimates depicted in Table `r tabnr`. The likelihood was computed under both the manipulated effect estimates (i.e., $L_{manipulated}$) and the genuine data (i.e., $L_{genuine}$). Table 4 indicates which cell sizes the various $n_{XX}$ refer to within the (simulated) data. After computing the likelihoods, we compared them to determine whether the simulated data were more likely to arise from the genuine trials ($L_{genuine}>L_{manipulated}$) or from the manipulated trials ($L_{manipulated}>L_{genuine}$). Note that comparing the likelihoods is a minor deviation from the preregistration, where we initially planned on using $p$-value comparisons ([osf.io/vnmzc](https://osf.io/vnmzc)).
